Higher-Order unification for free!

### 2024-04-12 paper/main.tex
sec 2.2 with spellchecker

In order to specify unification we need to
define the equational theory and
substitution (unification-variable assignment).

\subsubsection{Term equality: \Eo and \Ee}
For both languages we extend the equational theory
over ground terms to the full language by adding the reflexivity of
unification variables (a variable is equal to itself).

The first four rules are common to both equalities
and define the usual congruence over terms. Since
we use an HOAS encoding they also capture $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

@@

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app [F|L]}>>.
Note that the symbol \elpiIn{|} separates the head of a list from the tail.
\todo{explain forall2}

@@

We write $\sigma = \{~ X \mapsto t ~\}$ for the substitution that assigns
the term $t$ to the variable $X$.
We write $\sigma t$ for the application of
the substitution to a term $t$, and $\sigma X = \{~ \sigma t ~|~ t \in X ~\}$ when
$X$ is a set of terms.
We write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$.
% The domain of a substitution is the set of unification variables for which
% it provides an assignment.
% We write $\sigma \cup \sigma'$ to denote the concatenation of
% two substitutions whose domains are disjoint.
We shall use $\rho$ for \Fo{} substitutions,
and $\sigma$ for the \Ho ones.
For brevity, in this section we consider
the substitution for \Fo{} and \Ho{} identical.
We defer to \cref{sec:grounwork} a more precise description
pointing out theirs differences.

@@

\noindent
We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
reflected by signature above that relates two substitutions.
We write $t_1 \Ue t_2 \mapsto \sigma'$ when
the initial substitution $\sigma$ is empty.
% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
% the domains of $\sigma$ and $\sigma'$ are disjoint.
We write \llambda as the set of terms that are in the pattern-fragment, i.e.
every higher-order variable is applied to a list of distinct names.

>>>>

In order to specify unification, we need to
define the equational theory and
substitution (unification-variable assignment).

\subsubsection{Term equality: \Eo and \Ee}
For both languages, we extend the equational theory
over ground terms to the full language by adding the reflexivity for
unification variables (a variable is equal to itself).

The first four rules are common to both equalities
and define the usual congruence over terms. Since
we use an HOAS encoding, they also capture $\alpha$-equivalence.
In addition to that, \Eo has rules for $\eta$ and $\beta$-equivalence.

@@

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
For brevity, we omit the code of \elpiIn{beta}:
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app [F|L]}>>.
Note that the symbol \elpiIn{|} separates the head of a list from the tail.
\todo{explain forall2}

@@

We write $\sigma = \{~ X \mapsto t ~\}$ for the substitution that assigns
the term $t$ to the variable $X$.
We write $\sigma t$ for the application of
the substitution to a term $t$, and $\sigma X = \{~ \sigma t ~|~ t \in X ~\}$ when
$X$ is a set of terms.
We write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$.
% The domain of a substitution is the set of unification variables for which
% it provides an assignment.
% We write $\sigma \cup \sigma'$ to denote the concatenation of
% two substitutions whose domains are disjoint.
We shall use $\rho$ for \Fo{} substitutions,
and $\sigma$ for the \Ho ones.
For brevity, in this section, we consider
the substitution for \Fo{} and \Ho{} identical.
We defer to \cref{sec:grounwork} a more precise description
pointing out their differences.

@@

\noindent
We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
Note that $\sigma'$ is a refined (i.e. extended) version of $\sigma$; this is
reflected by the signature above that relates two substitutions.
We write $t_1 \Ue t_2 \mapsto \sigma'$ when
the initial substitution $\sigma$ is empty.
% Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
% the domains of $\sigma$ and $\sigma'$ are disjoint.
We write \llambda as the set of terms that are in the pattern-fragment, i.e.
every higher-order variable is applied to a list of distinct names.
